---
title: "Report 2a: Supervised Learning"
author:
- Yunhao Bai
- Zhiyi Chen
- Michelle Guan
- Huiqiong Wu
date: "2 March, 2022"
output:
  pdf_document:
    number_sections: no
    toc_depth: 4
  word_document:
    toc_depth: '4'
  html_document:
    df_print: paged
    toc_depth: '4'
subtitle: 'BUS 212A: Big Data II'
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhf{}
- \rhead{}
- \lhead{}
- \cfoot{\thepage}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(tidyverse)
library(leaps)
library(forecast)
library(car)
library(caret)
library(psych)
library(GGally)
library(gplots)
library(referenceIntervals)
```
\centering
\raggedright
\newpage
\tableofcontents

\newpage

### Data Cleaning
```{r, comment=NA}
setwd("/Users/chenzhiyi/Desktop/BUS 212A Project")
housing<-read.csv("new_housing.csv")
housing <- subset(housing, select = -c(1:3, 5, 17:18, 20))
#characterize variables
housing$cats_allowed = as.character(housing$cats_allowed)
housing$dogs_allowed = as.character(housing$dogs_allowed)
housing$smoking_allowed = as.character(housing$smoking_allowed)
housing$wheelchair_access = as.character(housing$wheelchair_access)
housing$electric_vehicle_charge = as.character(housing$electric_vehicle_charge)
housing$comes_furnished = as.character(housing$comes_furnished)
#multiple regression with price as target
lm_housing <- lm(price~. ,data=housing)
options(scipen = 999)
#summary(lm_housing)
vif(lm_housing)
```

### Descriptive Analysis

```{r, comment= NA}
median(housing$price)
mode(housing$laundry_options)
summary(is.na(housing$parking_options))

```

Descriptive analysis
Target variable: price_range (categorical)
    [Low; High]
    integer replacement: [0, 1]
Target variable: price (numeric)
Other variables (predictor variables):
 1. sqfeet: square feet (numeric)
 2. beds: number of beds (numeric)
 3. baths: number of baths (numeric)
 4. cats_allowed: (binary)
 5. dogs_allowed: (binary)
 6. smoking_allowed: (binary)
 7. wheelchair_access: (binary)
 8. electric_vehicle_charge: (binary)
 9. comes_furnished: (binary)
 10. laundry_options: (categorical)
        [no laundry on site, laundry in bldg, laundry on site,  w/d hookups, w/d in unit] 
        integer replacement: [1:4]
 11. parking_options: (categorical)
        [no parking, carpot, attached garage,  off-street parking, street parking, valet parking, detached garage]
        integer replacement: [1:6]
 12. state: (categorical) {for linear model}
 
The dataset structure: more categorical data than numeric data

```{r, comment=NA}
housingP<-housing
housingPR <- housing
housingPR <- housingPR %>%
 mutate(price_range = ifelse(
    price %in% c(0:1125), 0,
    ifelse(price %in% c(1126:20000000),
    1, "NA"))
 )
housingP$price_range <- housingPR$price_range
```

```{r, commeng=NA}
describe(housingP$price)
describe(housingP$sqfeet)
describe(housingP$beds)
describe(housingP$baths)
summary(housingP$price)
summary(housingP$sqfeet)
summary(housingP$beds)
summary(housingP$baths)
getmode <- function(x) {
  uniqx <- unique(x)
  uniqx[which.max(tabulate(match(x,uniqx)))]
}
x <- c(housingP$price)
result1 <- getmode(x)
print(result1)

y <- c(housingP$sqfeet)
result2 <- getmode(y)
print(result2)

z <- c(housingP$beds)
result3 <- getmode(z)
print(result3)

i <- c(housingP$baths)
result4 <- getmode(i)
print(result4)

ggplot(housingP, aes(x=price))+  # not normal, skewed, but sample size here is large enough
  geom_histogram()+
  ggtitle("Distribution of rental price")

ggplot(housingP, aes(x=beds)) +
  geom_histogram()+
  ggtitle("Distribution of number of beds")

ggplot(housingP, aes(x=sqfeet)) +  # numeric data's distribution and make it more Gaussian
  geom_histogram()+
  ggtitle("Distribution of square feet")

ggplot(housingP, aes(x=baths)) +  # numeric data's distribution and make it more Gaussian
  geom_histogram()+
  ggtitle("Distribution of number of baths")
```

#### Numeric data

```{r}
ggplot(housingP, aes(x=price_range, y=price, color=price_range))+
  geom_boxplot() + 
    xlab("Price range") +
    ylab("Rental price") +
    facet_wrap("price_range", scales = "free")+
    ggtitle("Box Plot:Price Range by Rental Price")

ggplot(housingP, aes(x=price_range, y=sqfeet, color=price_range))+
  geom_boxplot() + 
    xlab("Price range") +
    ylab("Square feet") +
    facet_wrap("price_range", scales = "free")+
    ggtitle("Box Plot:Price Range by Square Feet")

ggplot(housingP, aes(x=price_range, y=beds, fill=price_range)) +
    geom_violin() +
    xlab("Price range") +
    ylab("Number of beds") +
    facet_wrap("price_range", scales = "free") +
    ggtitle("Violin Plot:Price Range by Number of Beds")

ggplot(housingP, aes(x=price_range, y=baths, color=price_range))+
  geom_violin() + 
    xlab("Price range") +
    ylab("Number of baths") +
    facet_wrap("price_range", scales = "free")+
    ggtitle("Violin Plot:Price Range by Number of Baths")

ggplot(housingP, aes(x=beds, fill=price_range))+
  geom_bar()+
  ylab("Number of beds") +
  facet_wrap("price_range", scales = "free") +
  ggtitle("Distribution of Number of Beds by Price Range")

ggplot(housingP, aes(x=baths, fill=price_range))+
  geom_bar()+
  ylab("Number of baths") +
  facet_wrap("price_range", scales = "free") +
  ggtitle("Distribution of Number of Baths by Price Range")

# scatter plor matrix
housing.num <- c("price", "sqfeet", "beds", "baths", "price_range")
housing.numeric <- housingP[housing.num]
typeof(housing.numeric$price_range)
housing.numeric$price_range <- as.numeric(as.character(housing.numeric$price_range))
heatmap.2(cor(housing.numeric), dendrogram = "none",cellnote = round(cor(housing.numeric), 2), notecol = "black", key = FALSE, trace = 'none', margins = c(10,10)) 
ggpairs(housing.numeric)
```

#### Categorical Data

```{r}
ggplot(housingP, aes(x=cats_allowed, y=price, fill=cats_allowed)) +
  geom_boxplot()+
  facet_wrap("cats_allowed", scales = "free") 
ggplot(housingP, aes(x=dogs_allowed, y=price, fill=dogs_allowed)) +
  geom_boxplot()+
  facet_wrap("dogs_allowed", scales = "free")
ggplot(housingP, aes(x=smoking_allowed, y=price, fill=smoking_allowed)) +
  geom_boxplot()+
  facet_wrap("smoking_allowed", scales = "free")
ggplot(housingP, aes(x=wheelchair_access, y=price, fill=wheelchair_access)) +
  geom_boxplot()+
  facet_wrap("wheelchair_access", scales = "free")
ggplot(housingP, aes(x=electric_vehicle_charge, y=price, fill=electric_vehicle_charge)) +
  geom_boxplot()+
  facet_wrap("electric_vehicle_charge", scales = "free")
ggplot(housingP, aes(x=comes_furnished, y=price, fill=comes_furnished)) +
  geom_boxplot()+
  facet_wrap("comes_furnished", scales = "free")
ggplot(housingP, aes(x=laundry_options, y=price, fill=laundry_options)) +
  geom_boxplot()+
  facet_wrap("laundry_options", scales = "free_x")
ggplot(housingP, aes(x=parking_options, y=price, fill=parking_options)) +
  geom_boxplot()+
  facet_wrap("parking_options", scales = "free_x")
```

In matrix graph, information can obtain from the matrix graph: 
  1. potential curvilinear relationship: beds, sqfeet, baths [interaction term/ polynomial term for later model building]
  2. linear relationship

```{r}
ggplot(housingP, aes(x=price_range, fill = price_range)) +
  geom_bar()
# check for curviliearity (sqfeet and price)
ggplot(housingP, aes(x=sqfeet, y=price))+
  geom_point()+
  stat_smooth() +
  ggtitle("Relationship between price and sqfeet")

```

### Multiple Regression
### Diagnostic Plot

```{r, comment=NA}
par(mfrow=c(2,2))
plot(lm_housing)
```

#### Histogram of Residuals

```{r, comment=NA}
ggplot(data = housing, aes(x = lm_housing$residuals)) +
    geom_histogram(fill = 'steelblue', color = 'black') +
    labs(title = 'Histogram of Residuals', x = 'Residuals', y = 'Frequency')
```

#### Remove Outliers
##### Interquartile range

```{r}
housing$price = log(housing$price)
q1 <- quantile(housing$price, 0.25, na.rm=TRUE)
q3 <- quantile(housing$price, 0.75, na.rm=TRUE)
iqr <- IQR(housing$price, na.rm=TRUE)

clean_housing <- subset(housing, housing$price > (q1-1.5*iqr) & housing$price < (q3+1.5*iqr))

lm_housing <- lm(price~. ,data=clean_housing)
options(scipen = 999)
#summary(lm_housing)
vif(lm_housing)
par(mfrow=c(2,2))
plot(lm_housing)

#count outliers
outliers<-count(housing)-count(clean_housing)
percent_outliers<-outliers/count(housing)
percent_outliers

```




#### Skewness 

```{r}

#sqfeet
hist(clean_housing$sqfeet)
#sqfeet is right-skewed
#change the sqfeet to log term
clean_housing$sqfeet = log10(clean_housing$sqfeet)
#nearly become guassian
hist(clean_housing$sqfeet)

```


### Train and Test

```{r}
set.seed(2)
train.index <- sample(c(1:dim(clean_housing)[1]), dim(clean_housing)[1]*0.6)  
housing.train <- clean_housing[train.index, ]
housing.valid <- clean_housing[-train.index, ]
lm_housing <- lm(price ~., data=housing.train)

```


### Selection

#### Forward Selection

```{r, comment=NA}
#forward selection
lm_housing_null <- lm(price~1, data=housing.valid)
housing.lm.step.forward <- step(lm_housing_null, list(upper=lm_housing), direction="forward")
#summary(housing.lm.step.forward)
housing.lm.step.forward.pred <- predict(housing.lm.step.forward, housing.valid)
accuracy(housing.lm.step.forward.pred, housing.valid$price)


#price ~ state + sqfeet + laundry_options + parking_options + cats_allowed + smoking_allowed + electric_vehicle_charge + baths + beds

```

#### Backward Elimination
```{r, comment=NA}
housing.lm.step.backward <- step(lm_housing, direction="backward")
#summary(housing.lm.step.backward)
housing.lm.step.backward.pred <- predict(housing.lm.step.backward, housing.valid)
accuracy(housing.lm.step.backward.pred, housing.valid$price)


#price ~ sqfeet + beds + baths + cats_allowed + dogs_allowed + smoking_allowed + electric_vehicle_charge + laundry_options + parking_options + state
```

#### Bidirection Search
```{r, comment=NA}
housing.lm.step.both <- step(lm_housing, direction="both")
#summary(housing.lm.step.both)
housing.lm.step.both.pred <- predict(housing.lm.step.both, housing.valid)
accuracy(housing.lm.step.both.pred, housing.valid$price)



#price ~ sqfeet + beds + baths + cats_allowed + dogs_allowed + smoking_allowed + electric_vehicle_charge + laundry_options + parking_options + state

```

#### Higher-order terms

```{r, comment=NA}
#use the logged-sqfeet term as polynomial 
lm.poly <- lm(price ~ sqfeet + I(sqfeet^2) + beds + baths + cats_allowed + dogs_allowed + smoking_allowed + wheelchair_access + electric_vehicle_charge + comes_furnished + laundry_options + parking_options + state, data=housing.train)
housing.lm.poly.pred <- predict(lm.poly, housing.valid)
summary(lm.poly)
```

#### Adjusted R-squared and RMSE

```{r}
library(caret)
df<- data.frame(AdjR2=c(summary(housing.lm.step.forward)$adj.r.squared,
                        summary(housing.lm.step.backward)$adj.r.squared,
                        summary(housing.lm.step.both)$adj.r.squared,
                        summary(lm.poly)$adj.r.squared), 
                RMSE = c(RMSE(housing.lm.step.forward.pred, as.numeric(housing.valid$price)),
                         RMSE(housing.lm.step.backward.pred,as.numeric(housing.valid$price)),
                         RMSE(housing.lm.step.both.pred, as.numeric(housing.valid$price)),
                         RMSE(housing.lm.poly.pred, as.numeric(housing.valid$price)))
                )

rownames(df)<-c("Forward Selection", "Backward Elimination", "Bidirectional Search", "Polynomial Regression")
print(df)
# forward selection has higher adjr2 and lower RMSE
# forward selection is the best model
```
#### Best Model

```{r, comment=NA}
#best
#summary(housing.lm.step.forward)
```

### Logistic Regression

Logistic model
Multicollinearity affects the coefficients and p-values, but it does not influence the predictions, precision of the predictions, and the goodness-of-fit statistics. If your primary goal is to make predictions, and you don’t need to understand the role of each independent variable, you don’t need to reduce severe multicollinearity.

```{r}
# remove outliers by using IQR
q1 <- quantile(housingP$price, 0.25)
print(q1)
q3 <- quantile(housingP$price, 0.75)
print(q3)
iqr <- IQR(housingP$price)
print(iqr)
# from statistic features of variables shown, the distribution of our variables are skewed so it is better to IQR method to help remove outliars
# upper limit: q3+1.5*iqr
# lower limit: q1-1.5*iqr
clean_housing <- subset(housingP, housingP$price > (q1-1.5*iqr) & housingP$price < (q3+1.5*iqr))
iqr
# therefore, we get the percentage of outliers to be removed 
outliers<-count(housingP)-count(clean_housing)
percent_outliers<-outliers/count(housingP)
print(percent_outliers)

# data transformation --> more normalized

housing_clean_log <- log(clean_housing$price)
clean_housing$price <- housing_clean_log
clean_housing$sqfeet <- log(clean_housing$sqfeet)
```


```{r}
# replace categorical values with numbers
clean_housing$laundry_options <- case_when(
  clean_housing$laundry_options == "no laundry on site" ~ 0,
  clean_housing$laundry_options == "laundry in bldg" ~ 1,
  clean_housing$laundry_options == "laundry on site" ~ 2,
  clean_housing$laundry_options == "w/d hookups" ~ 3,
  clean_housing$laundry_options == "w/d in unit" ~ 4
)

clean_housing$parking_options <- case_when(
  clean_housing$parking_options == "no parking" ~ 0,
  clean_housing$parking_options == "carport" ~ 1,
  clean_housing$parking_options == "attached garage" ~ 2,
  clean_housing$parking_options == "off-street parking" ~ 3,
  clean_housing$parking_options == "street parking" ~ 4,
  clean_housing$parking_options == "valet parking" ~ 5,
  clean_housing$parking_options == "detached garage" ~ 6
)
clean_housing <- clean_housing[-c(1:3, 5,17, 18,19,20)]
clean_housing$price_range <- as.numeric(as.character(clean_housing$price_range))

set.seed(2)
train.index <- sample(c(1:dim(clean_housing)[1]), dim(clean_housing)[1]*0.6)  
train.df <- clean_housing[train.index, ]
valid.df <- clean_housing[-train.index, ]
```

```{r}
logit.reg_logit1 <- glm(price_range ~.-price-sqfeet+I(sqfeet^2), data = train.df, family = binomial(link = "logit"))
options(scipen=999)
summary(logit.reg_logit1)

VIFcheck <- lm( price_range ~.-price-sqfeet+I(sqfeet^2), data = train.df)
vif(VIFcheck)
```
```{r}
logit.reg_logit2 <- glm(price_range ~.-price-dogs_allowed-comes_furnished-sqfeet-wheelchair_access+I(sqfeet^2), data = train.df, family = binomial(link = "logit"))
options(scipen=999)
summary(logit.reg_logit2)

VIFcheck <- lm( price_range ~.-price-dogs_allowed-comes_furnished-sqfeet-wheelchair_access+I(sqfeet^2), data = train.df)
vif(VIFcheck)

par(mfrow=c(2,2)) 
plot(logit.reg_logit2)
```
```{r}
#remove extreme outlier
# from Normal Q-Q plot, we find data at 2837 row is outlier here
train.df <- train.df[-c(2837), ]
logit.reg_logit2 <- glm(price_range ~.-price-dogs_allowed-comes_furnished-sqfeet-wheelchair_access+I(sqfeet^2), data = train.df, family = binomial(link = "logit"))
options(scipen=999)
summary(logit.reg_logit2)

VIFcheck <- lm( price_range ~.-price-dogs_allowed-comes_furnished-sqfeet-wheelchair_access+I(sqfeet^2), data = train.df)
vif(VIFcheck)

par(mfrow=c(2,2)) 
plot(logit.reg_logit2)
```


```{r}
logit.reg.pred <- predict(logit.reg_logit2, valid.df, type = "response")
data.frame(actual = valid.df$price_range, predicted = logit.reg.pred)
library(gains)
gain <- gains(valid.df$price_range, logit.reg.pred, groups=10)
plot(c(0,gain$cume.pct.of.total*sum(valid.df$price_range))~c(0,gain$cume.obs), 
     xlab="# cases", ylab="Cumulative", main="", type="l")
lines(c(0,sum(valid.df$price_range))~c(0, dim(valid.df)[1]), lty=2)

```

```{r}
# odds
odds.df <- data.frame(odds = exp(coef(logit.reg_logit2))) 
vec.odds <- odds.df$odds - 1
odds.df$odds_1 <- vec.odds
odds.df
```

```{r}
summary(logit.reg.pred)
confusionMatrix(as.factor(ifelse(logit.reg.pred > 0.5, 1, 0)), as.factor(valid.df$price_range))
confusionMatrix(as.factor(ifelse(logit.reg_logit2$fitted.values > 0.5, 1, 0)), as.factor(train.df$price_range))
# F1 Score = 2 * (Precision * Recall) / (Precision + Recall)
F1 <- 2* (0.7230*0.6179) / (0.7230+0.6179)
F1
```

```{r}
residuals_logit <- c(logit.reg_logit2$residuals)
residuals_logit <- as.data.frame(residuals_logit)
summary(logit.reg_logit2$residuals)
zero <- logit.reg_logit2$residuals == 0
summary(zero)
ggplot(residuals_logit,aes(x=residuals_logit)) +
        geom_histogram()
```

```{r}
library(caret)
library(FNN)
library(class)
```

```{r}
plot(price ~ exp(sqfeet), data=train.df, pch=ifelse(train.df$price_range==0, "o", "+"))
```

### KNN

```{r}
train.df$price_range <- as.factor(train.df$price_range)
levels(train.df$price_range) = make.names(levels(factor(train.df$price_range)))
set.seed(2)

#- Define controls
x = trainControl(method = "repeatedcv",
                 number = 10,
                 repeats = 3,
                 classProbs = TRUE,
                 summaryFunction = twoClassSummary)

knn = train(price_range~. , data = train.df[, 2:13], method = "knn",
               preProcess = c("center","scale"),
               trControl = x,
               metric = "ROC",
               tuneLength = 10)
knn

train.df$predicted <- predict(knn, train.df, "prob")[,2]

train.df$predicted = ifelse(train.df$predicted > 0.5, 1,0)
train.df$price_range  = ifelse(train.df$price_range == 'X0', 0,1)
confusionMatrix(factor(train.df$predicted),
                factor(train.df$price_range))

valid.df$predicted <- predict(knn, valid.df, "prob")[,2]

valid.df$predicted = ifelse(valid.df$predicted > 0.5, 1,0)

confusionMatrix(factor(valid.df$predicted),
                factor(valid.df$price_range))


```


```{r}
F1_knn <- 2 * (0.7670*0.6312) / (0.7670+0.6312)
F1_knn
```









