---
title: "report2b"
output: html_document
---

```{r}
library(dclust)
library(caret)
library(class)
library(tidyverse)
library(dplyr)
library(factoextra)
rm(list = ls())
rental.df <- read.csv("/Users/chenzhiyi/Desktop/BUS 212A Project/iqr_price.csv")
# remove 'type' column 

#rental.df <- subset(rental.df, select = -c(2))
```

Rental prices of apartment in United States
Purpose of the clustering analysis:
  To gain insights from the clustering analysis of market structure and market segmentation based on the relevant variables in dataset
 1. sqfeet: square feet (numeric)
 2. beds: number of beds (numeric)
 3. baths: number of baths (numeric)
 4. cats_allowed: (binary)
 5. dogs_allowed: (binary)
 6. smoking_allowed: (binary)
 7. wheelchair_access: (binary)
 8. electric_vehicle_charge: (binary)
 9. comes_furnished: (binary)
 10. laundry_options: (categorical)
        [no laundry on site, laundry in bldg, laundry on site,  w/d hookups, w/d in unit] 
        numeric indicators: [1:4]
 11. parking_options: (categorical)
        [no parking, carpot, attached garage,  off-street parking, street parking, valet parking, detached garage]
        numeric indicator: [1:6]
 12. state: (categorical) {for linear model}
 13. price: (numeric)
 14. price range: (binary) 
        [High, Low]
        numeric indicator: [1, 2]
 
The dataset structure: 
 1. More categorical data than numeric data
 2. Low dimensional data: the number of observations are more than the number of features


1. Descriptive statistics: measures of central tendency and measures of variability (spread)
```{r}
#change to numeric
rental.df$price <- as.numeric(rental.df$price)
rental.df$sqfeet <- as.numeric(rental.df$sqfeet)
rental.df$beds <- as.numeric(rental.df$beds)
rental.df$baths <- as.numeric(rental.df$baths)

str(rental.df)
```




2. Distance:

```{r}
#elbow method
rental.df.norm <- sapply(rental.df[,c(1,2,3,4)], scale)

#fviz_nbclust(rental.df.norm, kmeans, method = "wss") +
#  geom_vline(xintercept = 5, linetype = 2) +
#  labs(subtitle = "Elbow method")
#suggesting k=5 is a good choice
```

```{r}
#K-means
set.seed(2)
rental.km <- kmeans(rental.df.norm, 5)#k=5
rental.km$centers
```


```{r}
d1 <- dist(rental.df.norm, method = "canberra")
d2 <- dist(rental.df.norm, method = "binary")#not use
d3 <- dist(rental.df.norm, method = "minkowski")
d4 <- dist(rental.df.norm, method = "euclidean")
d5 <- dist(rental.df.norm, method = "manhattan")
d6 <- dist(rental.df.norm, method = "maximum")

```

3.Hierarchical Cluster

```{r}
#canberra
hc1 <- hclust(d1, method = "ward.D")
plot(hc1, hang = -1, ann = FALSE)
memb1 <- cutree(hc1, h = 4000) #from elbow method
hist(memb1)
```

```{r}
#minkowski
hc3 <- hclust(d3, method = "ward.D")
plot(hc3, hang = -1, ann = FALSE)
memb3 <- cutree(hc3, h = 3000) #from elbow method
hist(memb3)
```

```{r}
#euclidean
hc4 <- hclust(d4, method = "ward.D")
plot(hc4, hang = -1, ann = FALSE)
memb4<- cutree(hc4, h = 4000) #from elbow method
hist(memb4)
```

```{r}
#manhattan
hc5 <- hclust(d5, method = "ward.D")
plot(hc5, hang = -1, ann = FALSE)
memb5 <- cutree(hc5, h = 5000) #from elbow method
hist(memb5)
#seems to be the best?
```

```{r}
#maximum
hc6 <- hclust(d6, method = "ward.D")
plot(hc6, hang = -1, ann = FALSE)
memb6 <- cutree(hc6, h = 2000) #from elbow method
hist(memb6)
```



```{r}
#combined to the original data 
#rental.km$cluster
hclust_clusters <- memb1
km_clusters <- rental.km$cluster
rental.df2 <- data.frame(rental.df, hclust_clusters, km_clusters)
```

```{r}
plot(c(0), xaxt = 'n', ylab = "", type = "l", 
     ylim = c(min(rental.km$centers), max(rental.km$centers)), xlim = c(0, 5))

# label x-axes
axis(1, at = c(1:4), labels = c("price", "sqfeet", "beds", "baths"))

# plot centroids
for (i in c(1:5))
  lines(rental.km$centers[i,], lty = i, lwd = 2, col = switch(i, "black", "red", 
                                                       "green", "purple", "yellow"))
                                                       
                                                       
# name clusters
text(x = 0.5, y = rental.km$centers[, 1], labels = paste("Cluster", c(1:5)))

```

```{r}
heatmap(as.matrix(rental.df.norm), Colv = NA, hclustfun = hclust, 
        col=rev(paste("grey",1:99,sep="")))

```

```{r}
#hierachical cluster
center = aggregate(rental.df.norm, list(cluster = hclust_clusters), mean)
hclust_center<-center[,-1]

plot(c(0), xaxt = 'n', ylab = "", type = "l", 
     ylim = c(min(hclust_center), max(hclust_center)), xlim = c(0, 5))
axis(1, at = c(1:4), labels = c("price", "sqfeet", "beds", "baths"))
for (i in c(1:3))
  lines(unlist(hclust_center[i,], use.names=FALSE), lty = i, lwd = 2, col = switch(i, "black", "red", 
                                                       "green", "purple", "yellow"))
text(x = 0.5, y = hclust_center[, 1], labels = paste("Cluster", c(1:3)))

```

```{r}
library(grid)
library(gridExtra)
rental.df2$hclust_clusters<-as.factor(rental.df2$hclust_clusters)
rental.df2$km_clusters<-as.factor(rental.df2$km_clusters)
p1 <- ggplot(rental.df2, aes(x=price, y=sqfeet, color=hclust_clusters)) + 
  geom_point() + 
  labs(x="Dim1",
       y="Dim2",
       title="Hierarchical Cluster") +
  theme_classic() +
  theme(plot.title = element_text(size=10),
        axis.title = element_text(size=10),
        legend.title = element_text(size=10)) +
  theme(legend.position = "none")

p2 <- ggplot(rental.df2, aes(x=price, y=sqfeet, color=km_clusters)) + 
  geom_point() + 
  labs(x="Dim1",
       y="Dim2",
       title="KMeans Cluster") +
  theme_classic() +
  theme(plot.title = element_text(size=10),
        axis.title = element_text(size=10),
        legend.title = element_text(size=10)) +
  theme(legend.position = "none")
grid.arrange(p1, p2, ncol=2, nrow=1, top=textGrob("Hierarchical vs Kmeans Clustering") )
```


```{r}
#write.csv(rental.df2,"/Users/chenzhiyi/Desktop/BUS 212A Project/data_clusters.csv", row.names = FALSE)
```




